Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Responses to Ben's Sec 3 edits

The first step of most data analysis is "cleaning" the raw data so it can be explored. Before data can be explored, they must be collected. Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  \ben{In this setting, c}ompartment totals (amounts of individuals in each state) are then imputed from those case counts \ben{and using} other information about the disease and the population of interest\footnote{\ben{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. \footnote{\ben{[Ben says: I'd like to replace these two sentences will a new one. The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best). Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g. subpopulations defined by age or sex).]}}There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.

@@

 In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.  An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways. \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there are three unique epidemiological states (like in the SIR model). Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}. The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.  Visualization tools for three-state models were developed because \ben{\sout{(1)}} SIR models are some of the most common and basic epidemic state-based models and \ben{\sout{(2)}}\footnote{\ben{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}} our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation. \ben{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\ben{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}} While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  \ben{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\ben{[Ben says: see edit below as well. This sentence seemed misplaced, and we have a very similar sentence a few lines later. I'm happy to be overruled here.]}}  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  \ben{\sout{In turn, this allows for} Both of these functions allow} users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

\shannon{[[NEW TEXT}
If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format. \ben{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\ben{[Ben says: this really isn't true anymore - that's more the next section.]}} In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output\footnote{\ben{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}} from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

Modeling and simulation output can be very complex objects, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.  The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to handle. To this end, \pkg{EpiCompare} provides a series of fortify-style methods, called \code{fortify_aggregate()} which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.  These fortify functions have output that is consistent with that of \code{agents_to_aggregate()}.  These standardized outputs can then be piped to summaries, tables, and plots.

Because \ben{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\ben{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}} we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations. These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
\shannon{]]}

@@

Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.

>>>>

The first step of most data analysis is "cleaning" the raw data so it can be explored. Before data can be explored, they must be collected. Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  In this setting, compartment totals (amounts of individuals in each state) are then imputed from those case counts and using other information about the disease and the population of interest.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. \footnote{\ben{[Ben says: I'd like to replace these two sentences will a new one. The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best). Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregati\shannon{on}, conditional on subpopulations (e.g. subpopulations defined by age or sex).]}\shannon{I like your edits}}There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`.\sout{, the tidyverse approach to the split-apply-combine paradigm for data analysis} \citep{wickham2011}.

@@

 In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.  An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways. \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there are three unique epidemiological states (like in the SIR model). Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}. The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.  Visualization tools for three-state models were developed because \ben{\sout{(1)}} SIR models are some of the most common and basic epidemic state-based models and \ben{\sout{(2)}}\footnote{\ben{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}\shannon{accepted}} our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation. \ben{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\ben{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}\shannon{accept}} While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  \ben{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\ben{[Ben says: see edit below as well. This sentence seemed misplaced, and we have a very similar sentence a few lines later. I'm happy to be overruled here.]}\shannon{accept, yeah it's unnecessary}}  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  Both of these functions allow users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

\shannon{[[NEW TEXT}
If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format. \ben{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\ben{[Ben says: this really isn't true anymore - that's more the next section.]}\shannon{accept}} In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model \shannon{objects}\ sout{output} \footnote{\ben{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}\shannon{trying out objects..}} from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

Modeling and simulation \shannon{objects} \sout{output} can be very complex objects, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.  The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to handle. To this end, \pkg{EpiCompare} provides a series of fortify-style methods, called \code{fortify_aggregate()} which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.  These fortify functions have output that is consistent with that of \code{agents_to_aggregate()}.  These standardized outputs can then be piped to summaries, tables, and plots.

Because \ben{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\ben{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]} \shannon{accept}} we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations. These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
\shannon{]]}

@@

\shannon{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.  We first talk about the step in the pipeline and then how epicompare fits into that pipeline.  The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.  I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}

Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare \shannon{and assess} different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one season or cycle of an epidemic has occurred.

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.tex
Responses to Ben's Sec 3 edits

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. \textcolor{violet}{In this setting, c}ompartment totals
(amounts of individuals in each state) are then imputed from those case
counts \textcolor{violet}{and using} other information about the disease
and the population of
interest\footnote{\textcolor{violet}{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.
In \pkg{EpiCompare}, we focus on understanding the overall impact of an
outbreak at the aggregate/population level, which allows for streamlined
examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
\footnote{\textcolor{violet}{[Ben says: I'd like to replace these two sentences will a new one. The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best). Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g. subpopulations defined by age or sex).]}}There
are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for differing disease dynamics in different subpopulations.
In \pkg{EpiCompare}, we facilitate subpopulation
analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
by extending \texttt{agents\_to\_aggregate()} to work with
\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
split-apply-combine paradigm for data analysis \citep{wickham2011}.

@@

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there are three unique epidemiological
states (like in the SIR model). Building on \pkg{ggplot2} and
\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
provides a way to explore how different subpopulations experience of an
epidemic \citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because \textcolor{violet}{\sout{(1)}} SIR models
are some of the most common and basic epidemic state-based models and
\textcolor{violet}{\sout{(2)}}\footnote{\textcolor{violet}{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}}
our three-dimensional simplex representation of these epidemics
emphasizes a time-invariant representation of the data (for a refresher
see Section \ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is often model fitting
and/or simulation.
\textcolor{violet}{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\textcolor{violet}{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}}
While \pkg{EpiCompare} does not focus on fitting models to data, we do
provide some flexible functions for simulation of basic discrete-time
epidemic-state models. These functions simulate individual-level
information based on practitioner estimated transition rates between
states and can be combined with \texttt{agents\_to\_aggregate()} to view
these simulations through an aggregate lens. The function
\texttt{simulate\_SIR\_agents()} simulates a basic SIR epidemic with
user inputs for the number of simulations, the initial number in each
state, the infection and recovery parameters \((\beta, \gamma\)), and
the total number of discrete time steps.
\textcolor{violet}{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\textcolor{violet}{[Ben says: see edit below as well. This sentence seemed misplaced, and we have a very similar sentence a few lines later. I'm happy to be overruled here.]}}
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified state-transition matrix and other epidemic
parameters to allow the user to create simulations for an outbreak with
\textit{any} number of states and any number of transitions among them.
This flexibility in states can be used to also reflect group-based
dynamics.
\textcolor{violet}{\sout{In turn, this allows for} Both of these functions allow}
users to explore the space of models in an intuitive way without getting
bogged down by too much mathematical detail. For consistency, we have
made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.

@@

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format.
\textcolor{violet}{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\textcolor{violet}{[Ben says: this really isn't true anymore - that's more the next section.]}}
In \pkg{EpiCompare}, we provide (1) functions to standardize simulation
and model
output\footnote{\textcolor{violet}{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}}
from external packages and (2) a function to transform standardized
simulation and model output into a format amenable to time-invariant
analysis.\footnote{\textcolor{orange}{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

Modeling and simulation output can be very complex objects, and as a
result, a number of epidemic modeling \proglang{R} packages return a
special class. The special classes often contain a plethora of
information about residuals, model diagnostics, input parameters, and
more. While incredibly useful, these special classes can be difficult
for novice coders to handle. To this end, \pkg{EpiCompare} provides a
series of fortify-style methods, called \code{fortify_aggregate()} which
transform output from infectious disease modeling and simulation
packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
which contain information about the total number of individuals in each
state at a given time, for a given simulation. These fortify functions
have output that is consistent with that of
\code{agents_to_aggregate()}. These standardized outputs can then be
piped to summaries, tables, and plots.

Because
\textcolor{violet}{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\textcolor{violet}{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}}
we provide the function, \code{filament_compression()}, to transform
temporally defined epidemics to their filamental representations. These
filaments can then be fairly compared to one another or passed to
further time-invariant analysis tools described below.
\textcolor{orange}{]]}

@@

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare different models, and
simulations to one
another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
In \pkg{EpiCompare} we provide a set of comparison and assessment tools
for model and simulation results that extend beyond the standard
performance metrics (e.g.~mean squared error or AIC) and into the lens
of time-invariant
analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
We have found these tools to be specifically applicable for situations
where only one ``season'' or ``cycle'' of an epidemic has occurred.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing pseudo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated pseudo-density & proportion of simulations with lower estimated pseudo-density\\
\midrule
Simple SIR & 0.0036733 & 0\\
EpiModel SIR & 0.0028813 & 0\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. In this setting, compartment totals (amounts of individuals in
each state) are then imputed from those case counts and using other
information about the disease and the population of interest. In
\pkg{EpiCompare}, we focus on understanding the overall impact of an
outbreak at the aggregate/population level, which allows for streamlined
examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
\footnote{\textcolor{violet}{[Ben says: I'd like to replace these two sentences will a new one. The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best). Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregati\textcolor{orange}{on}, conditional on subpopulations (e.g. subpopulations defined by age or sex).]}\textcolor{orange}{I like your edits}}There
are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for differing disease dynamics in different subpopulations.
In \pkg{EpiCompare}, we facilitate subpopulation
analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
by extending \texttt{agents\_to\_aggregate()} to work with
\pkg{dplyr}::\texttt{group\_by()}.\sout{, the tidyverse approach to the split-apply-combine paradigm for data analysis}
\citep{wickham2011}.

@@

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there are three unique epidemiological
states (like in the SIR model). Building on \pkg{ggplot2} and
\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
provides a way to explore how different subpopulations experience of an
epidemic \citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because \textcolor{violet}{\sout{(1)}} SIR models
are some of the most common and basic epidemic state-based models and
\textcolor{violet}{\sout{(2)}}\footnote{\textcolor{violet}{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}\textcolor{orange}{accepted}}
our three-dimensional simplex representation of these epidemics
emphasizes a time-invariant representation of the data (for a refresher
see Section \ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is often model fitting
and/or simulation.
\textcolor{violet}{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\textcolor{violet}{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}\textcolor{orange}{accept}}
While \pkg{EpiCompare} does not focus on fitting models to data, we do
provide some flexible functions for simulation of basic discrete-time
epidemic-state models. These functions simulate individual-level
information based on practitioner estimated transition rates between
states and can be combined with \texttt{agents\_to\_aggregate()} to view
these simulations through an aggregate lens. The function
\texttt{simulate\_SIR\_agents()} simulates a basic SIR epidemic with
user inputs for the number of simulations, the initial number in each
state, the infection and recovery parameters \((\beta, \gamma\)), and
the total number of discrete time steps.
\textcolor{violet}{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\textcolor{violet}{[Ben says: see edit below as well. This sentence seemed misplaced, and we have a very similar sentence a few lines later. I'm happy to be overruled here.]}\textcolor{orange}{accept, yeah it's unnecessary}}
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified state-transition matrix and other epidemic
parameters to allow the user to create simulations for an outbreak with
\textit{any} number of states and any number of transitions among them.
This flexibility in states can be used to also reflect group-based
dynamics. Both of these functions allow users to explore the space of
models in an intuitive way without getting bogged down by too much
mathematical detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.

@@

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format.
\textcolor{violet}{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\textcolor{violet}{[Ben says: this really isn't true anymore - that's more the next section.]}\textcolor{orange}{accept}}
In \pkg{EpiCompare}, we provide (1) functions to standardize simulation
and model \textcolor{orange}{objects}~sout\{output\}
\footnote{\textcolor{violet}{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}\textcolor{orange}{trying out objects..}}
from external packages and (2) a function to transform standardized
simulation and model output into a format amenable to time-invariant
analysis.\footnote{\textcolor{orange}{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

Modeling and simulation \textcolor{orange}{objects} \sout{output} can be
very complex objects, and as a result, a number of epidemic modeling
\proglang{R} packages return a special class. The special classes often
contain a plethora of information about residuals, model diagnostics,
input parameters, and more. While incredibly useful, these special
classes can be difficult for novice coders to handle. To this end,
\pkg{EpiCompare} provides a series of fortify-style methods, called
\code{fortify_aggregate()} which transform output from infectious
disease modeling and simulation packages like \pkg{pomp} and
\pkg{EpiModel} into tidy-styled data frames which contain information
about the total number of individuals in each state at a given time, for
a given simulation. These fortify functions have output that is
consistent with that of \code{agents_to_aggregate()}. These standardized
outputs can then be piped to summaries, tables, and plots.

Because
\textcolor{violet}{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\textcolor{violet}{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]} \textcolor{orange}{accept}}
we provide the function, \code{filament_compression()}, to transform
temporally defined epidemics to their filamental representations. These
filaments can then be fairly compared to one another or passed to
further time-invariant analysis tools described below.
\textcolor{orange}{]]}

@@

\textcolor{orange}{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.  We first talk about the step in the pipeline and then how epicompare fits into that pipeline.  The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.  I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare
\textcolor{orange}{and assess} different models, and simulations to one
another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
In \pkg{EpiCompare} we provide a set of comparison and assessment tools
for model and simulation results that extend beyond the standard
performance metrics (e.g.~mean squared error or AIC) and into the lens
of time-invariant
analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
We have found these tools to be specifically applicable for situations
where only one season or cycle of an epidemic has occurred.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing pseudo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated pseudo-density & proportion of simulations with lower estimated pseudo-density\\
\midrule
Simple SIR & 0.0036733 & 0.00\\
EpiModel SIR & 0.0149686 & 0.02\\
\bottomrule
\end{tabular}
\end{table}
